Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-06-30 Manuscript/manuscript.Rmd
Edits through Introduction

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. However, a key limitation among most of these "TAD(loop)-callers" is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD(loop)-callers is their disregard for the fundamental roles of functional genomic elements on chromatin folding. It has been shown that particular genomic elements are enriched at TAD(loop) boundaries. Namely, the insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) under a supervised machine learning framework. Our method builds upon an optimized random forest classification model (RF) that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates as base pairs on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome.

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD(loop)-caller. Ideally we would perform simulation studies to benchmark *preciseTAD* against simulated boundaries. While there are existing methods for simulating Hi-C data sets [@forcato2017comparison; @zheng2020freehi], of which boundaries can be extracted from, to date, there are no such methods for simulating ChIP-seq data. However, given *preciseTAD* is applied on chromosome-specific sections of the linear genome, and on multiple cell lines, we feel that aggregating results across chromosomes provides strong empirical evidence of the superiority of *preciseTAD*. Another limitation is that *preciseTAD* does not provide hierarchical boundaries, but instead simply provides the bp coordinate that demarcates TADs(loops). Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries.

>>>>

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. [??? Check refs, the current ones are suboptimal. Check SpectralTAD paper for refs about TAD callers, use https://github.com/mdozmorov/HiC_tools#loop-callers for loop callers] However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. [??? Rewrite, simplify, describe enrichment first, loop extrusion second] Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) under a supervised machine learning framework. Our method builds upon an optimized random forest classification model (RF) that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates as base pairs on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability]

@@

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD caller. Ideally we would perform simulation studies to benchmark *preciseTAD* against simulated boundaries. While there are existing methods for simulating Hi-C data sets [@forcato2017comparison; @zheng2020freehi], of which boundaries can be extracted from, to date, there are no such methods for simulating ChIP-seq data. However, given *preciseTAD* is applied on chromosome-specific sections of the linear genome, and on multiple cell lines, we feel that aggregating results across chromosomes provides strong empirical evidence of the superiority of *preciseTAD*. Another limitation is that *preciseTAD* does not provide hierarchical boundaries, but instead simply provides the bp coordinate that demarcates TADs(loops). Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries.
